---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ConNIS <img src="./man/figures/logo.svg" alt="ConNIS" align="right" width="120"/>

<!-- badges: start -->
<!-- badges: end -->

The `ConNIS` package provides multiple methods for determining essential genes based on insertion sites of *TraDIS* data. It also supports the data driven determination of tuning/threshold value selection by the implementation of a subsample based instability approach. Currently the follwing methods are implemented:

* the *Consecutive Non-Insertion Sites* (ConNIS) method (Hanke et al. (2025))
* an implementation of the *Binomial* approach  of the `TSAS 2.0`  package with an additional weigh parameter
* an implementation of the *Tn5Gaps* method of the `TRANSIT` package with an additional weigh parameter
* the determination of essential genes by applying the geometric distribution


## Installation

You can install the development version of ConNIS from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("moritz-hanke/ConNIS")
```

## Example

We use a truncated real world dataset and a list of truncated insertion sites as an example for applying ConNIS.

```{r exampleRunConNIS}

library(ConNIS)

# Use the E. coli BW 25113 dataset but only the first 100 genes
truncated_ecoli <- ecoli_bw25113[50:100,]

# load the insertion sites by Goodall, 2018, but omit all insertion sites that are
# within the truncated_ecoli
truncated_is_pos <- is_pos[is_pos <= max(truncated_ecoli$end) &
                           is_pos >= min(truncated_ecoli$start)]

# run ConNIS with weight = 1
results_ConNIS <- 
  ConNIS(ins.positions = truncated_is_pos, 
       gene.names = truncated_ecoli$gene, 
       gene.starts = truncated_ecoli$start, 
       gene.stops = truncated_ecoli$end, 
       genome.length = max(truncated_ecoli$end), 
       weight = 1)

results_ConNIS
```

Using a simple "Bonferroni correction" with $\alpha=0.05$ for multiple testing problem ConNIS declared the follwing 13 genes as essential:

```{r exampleGetSignificantGenes}
results_ConNIS %>% filter(p_value <= 0.05/nrow(truncated_ecoli))
```

Next, we re-run ConNIS with a smaller weight and apply again a the "Bonferroni" method.

```{r rerunConNIS}
results_ConNIS <- 
  ConNIS(ins.positions = truncated_is_pos, 
       gene.names = truncated_ecoli$gene, 
       gene.starts = truncated_ecoli$start, 
       gene.stops = truncated_ecoli$end, 
       genome.length = max(truncated_ecoli$end), 
       weight = 0.2)

results_ConNIS %>% filter(p_value <= 0.05/nrow(truncated_ecoli))

```
Only 8 genes are declared essential since smaller weights will make it harder to label a gene (by chance) as 'essential'. 

Next, we give an example how to select a weight by the instability approach. We will use the parallel version of the function using `mclapply`. We use different weights. NOTE: We use only 30 subsamples for demonstration purpose. For real world applications we suggest $m \approx  500$ and parallel processing uization of the .

```{r instability}

# set weight
weights <- seq(0.2, 1, 0.2)

out <- 
  instabilities(
  method="ConNIS", 
  sig.level = 0.05, 
  p.adjust.mehtod = "bonferroni", 
  ins.positions = truncated_is_pos, 
  gene.names = truncated_ecoli$gene, 
  gene.starts = truncated_ecoli$start, 
  gene.stops = truncated_ecoli$end, 
  genome.length = max(truncated_ecoli$end), 
  weights = weights, 
  m = 30, 
  d = 0.5, 
  use.parallelization = T, 
  parallelization.type = "mclapply", 
  numCores = detectCores()-1, 
  seed = 1, 
  set.rng = "L'Ecuyer-CMRG")

out

```

You can also embed plots, for example:

```{r pressure, echo = FALSE}
#plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
