---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ConNIS <img src="./man/figures/logo.svg" alt="ConNIS" align="right" width="120"/>

<!-- badges: start -->
<!-- badges: end -->

The `ConNIS` package provides the implementation of the *Consecutive Non-Insertion Sites* (ConNIS) method (Hanke et al., 2025) which determines putative essential genes based on the probabilities to observe sequences of non-insertions within genes by chance. To set the weight value $w$ for adjusting the method for non-uniformaly distributed gene-wise insertion desists an instability approach is provided.

In addition, the following methods are implemented in the package:
* the *Binomial* approach  of the `TSAS 2.0`  package (Burger et al., 2017) with an additional weigh parameter
* the *Tn5Gaps* method of the `TRANSIT` package (DeJesus et al, 2015) with an additional weigh parameter
* the determination of essential genes by applying the *Geometric* distribution with an additional weight parameter


## Installation

You can install the development version of ConNIS from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("moritz-hanke/ConNIS")
```

## Example

We use a truncated real world dataset and a list of truncated insertion sites as an toy example for applying ConNIS.

```{r exampleRunConNIS, message=FALSE, warnings=FALSE}

library(ConNIS)

# Use the E. coli BW 25113 dataset but only the first 30 genes
truncated_ecoli <- ecoli_bw25113[1:30,]

# load the insertion sites by Goodall et al., 2018, but omit all insertion sites 
# that do not lie within the genomic region of truncated_ecoli
truncated_is_pos <- sort(is_pos[is_pos <= max(truncated_ecoli$end) &
                           is_pos >= min(truncated_ecoli$start)])

# run ConNIS with weight = 1
results_ConNIS <- 
  ConNIS(ins.positions = truncated_is_pos, 
       gene.names = truncated_ecoli$gene, 
       gene.starts = truncated_ecoli$start, 
       gene.stops = truncated_ecoli$end, 
       genome.length = max(truncated_ecoli$end), 
       weight = 1)

results_ConNIS
```

Using the "Bonferroni-Holm" correction with $\alpha=0.05$ for multiple testing adjustment ConNIS declared the follwing 12 genes as essential:

```{r exampleGetSignificantGenes}
results_ConNIS %>% filter(p.adjust(p_value, "BH") <= 0.05)
```

Next, we re-run ConNIS with a smaller weight value and apply again a the "Bonferroni" method.

```{r rerunConNIS}
results_ConNIS <- 
  ConNIS(ins.positions = truncated_is_pos, 
       gene.names = truncated_ecoli$gene, 
       gene.starts = truncated_ecoli$start, 
       gene.stops = truncated_ecoli$end, 
       genome.length = max(truncated_ecoli$end), 
       weight = 0.2)

results_ConNIS %>% filter(p.adjust(p_value, "BH") <= 0.05)

```
6 genes are declared essential since smaller weights will make it harder to label a gene (by chance) as 'essential'. 

Next, we give an example for selecting `weight` our of five different values by the instability approach. We will use the function's build-in parallelization option (`parallelization.type = "mclapply"`). NOTE: Only 10 subsamples are drawn for demonstration purpose. For real world applications we suggest $m \approx  500$.

```{r instability}

# set weight
weights <- seq(0.2, 1, 0.2)

instabilities_connis <- 
  instabilities(
  method="ConNIS", 
  sig.level = 0.05, 
  p.adjust.mehtod = "BH", 
  ins.positions = truncated_is_pos, 
  gene.names = truncated_ecoli$gene, 
  gene.starts = truncated_ecoli$start, 
  gene.stops = truncated_ecoli$end, 
  genome.length = max(truncated_ecoli$end), 
  weights = weights, 
  m = 10, 
  d = 0.5, 
  use.parallelization = T, 
  parallelization.type = "mclapply", 
  numCores = detectCores()-1, 
  seed = 1, 
  set.rng = "L'Ecuyer-CMRG")

instabilities_connis

```
Applying ConNIS again with the weight value that had the minimal instability, i.e. `weight = 0.6`, 7 genes are declared essential:

```{r}
results_ConNIS <- 
  ConNIS(ins.positions = truncated_is_pos, 
       gene.names = truncated_ecoli$gene, 
       gene.starts = truncated_ecoli$start, 
       gene.stops = truncated_ecoli$end, 
       genome.length = max(truncated_ecoli$end), 
       weight = 0.6)

results_ConNIS %>% filter(p.adjust(p_value, "BH") <= 0.05)
```


## References
* `Burger, B. T., Imam, S., Scarborough, M. J., Noguera, D. R. & Donohue, T. J. Combining genome-scale experimental and computational methods to identify essential genes in rhodobacter sphaeroides. mSystems 2 (2017). URL http://dx. doi.org/10.1128/msystems.00015-17`
* `DeJesus, M. A., Ambadipudi, C., Baker, R., Sassetti, C. & Ioerger, T. R. Transit - a software tool for himar1 tnseq analysis. PLOS Computational Biology 11, e1004401 (2015). URL http://dx.doi.org/10.1371/journal.pcbi.1004401`
* `Goodall, E. C. A. et al. The essential genome of escherichia coli k-12. mBio 9 (2018). URL http://dx.doi.org/10.1128/mBio.02096-17.`
